LEGWORK: A python package for computing the evolution and detectability \\ of stellar-origin gravitational-wave sources with space-based detectors

### 2021-11-12 src/ms.tex
need full path
update after SdM final comments

\title{LEGWORK: LISA Evolution and Gravitational Wave Orbit Kit}

@@

\begin{abstract}
    We present \lw{}, a Python package for making predictions about stellar-origin gravitational wave sources and their detectability in LISA and other space-based gravitational wave detectors. \lw{} can be used to evolve sources over time, calculate gravitational wave strains, compute signal-to-noise ratios and visualise the results. In this paper we give an overview of the structure and functionality of the package, as well as highlight several optimisations that make it particularly efficient for use in population studies which can contain tens-of-millions of sources. We demonstrate the various capabilities of \lw{} through a series of example use cases. These begin with a basic example of calculating the signal-to-noise ratio of a source but also include using \lw{} to calculate horizon distances and make comparisons between gravitational-wave detectors. We additionally explain in detail the derivation of the equations behind \lw{} in order to clarify some conflicts currently present in the literature.
\end{abstract}

\keywords{gravitational waves, LISA}

@@

The planned space-based gravitational wave detector LISA (Laser Interferometer Space Antenna) will present an entirely new view of gravitational waves by focusing on lower frequencies ($10^{-5} < f / \unit{Hz} < 10^{-1}$) than ground-based detectors. This will enable the study of many source classes that are undetectable with ground-based detectors including mergers of supermassive black holes that are undetectable with ground-based detectors \citep[e.g.][]{Begelman+1980, Klein+2016, Bellovary2019}, extreme mass ratio inspirals \citep[e.g.,][]{Berti2006, Barack2007, Babak2017, Moore2017}, and cosmological GW backgrounds \citep[e.g.,][]{Bartolo2016, Caprini2016, Caldwell2019}. However, this frequency regime is also of interest for the detection of local stellar-mass binaries during their inspiral phase. LISA is expected to detect Galactic stellar-origin binaries containing combinations of white dwarfs, neutron stars, and black holes, ranging from the numerous double white dwarf population, to the rare but loud double black hole population.

The potential to detect stellar-origin sources with LISA has been studied in the past \citep[e.g.][]{Nelemans+2001, Belczynski+2010}. More recently, the direct detection of gravitational waves with ground-based detectors has led to renewed interest in this topic with many recent papers addressing the issue \citep[e.g.][]{Christian+2017, Kremer+2017, Kremer+2018, Korol+2017, Korol+2018, Korol+2019, Korol+2020, Lamberts+2018, Lamberts+2019, Fang+2019, Andrews+2020, Lau+2020, Breivik+2020, Breivik+2020a, Roebber+2020, Chen+2020, Sesana+2020}.

@@

\lw{} is a Python package designed to streamline the process of making predictions of LISA detection rates for stellar-origin binaries such that it is as fast, reliable and simple as possible. This goal makes \lw{} unique among other implementations of gravitational-wave tools in the literature, which focus on a more broad coverage of the gravitational-wave spectrum and source classes, rather than an optimised approach for certain sources. \citep[e.g.][]{Moore+2015,Yi2021}. With \lw{} one can evolve the orbits of a binary or a collection of binaries and calculate their strain amplitudes for any range of frequency harmonics. One can compute the sensitivity curve for LISA or other future gravitational wave detectors (e.g. TianQin's curve, or that of a custom instrument) and use it to compute the signal-to-noise ratio of a collection of sources. Furthermore, \lw{} provides tools to visualise all of the results with easy-to-use plotting functions. Finally, \lw{} is fully tested to check for consistency in the derivations described below.

Specifically, we implement the expressions by \citet{Peters+1963} and \citet{Peters+1964} for the evolution of binary orbits due to the emission of gravitational waves, equations for the strain amplitudes and signal-to-noise ratios of binaries from various papers \citep[e.g.][]{Flanagan+1998, Finn+2000, Cornish2003, Barack+2004, Moore+2015} and approximations for the LISA and TianQin sensitivity curves given in \citet{Robson+2019} and \citet{Huang+2020} respectively.

@@

@@

@@

The \lw{} package is composed of seven modules that each focus on a particular aspect of calculations useful for LISA gravitational wave sources. In Fig.~\ref{fig:package_overview}, we illustrate the general structure of the package with each of its modules. The source module is the central module of the package and provides an simple interface to the functions in the rest of the modules. For more complex analyses, users may want to interact directly with individual modules, particularly those in the top row of Fig.~\ref{fig:package_overview} as they comprise the core functionality of \lw{}. Below we explain the capabilities of each module in detail.

@@

\lwModLink{strain} contains two functions that compute a binary's gravitational wave strain and characteristic strain amplitude respectively. Each of these functions is capable of computing the strain for a 3-dimensional array of binaries at any number of timesteps and evaluated at any number of frequency harmonics. We discuss the relevant equations in Sec.~\ref{sec:deriv_strain}.

\lwModLink{psd} is used for evaluating the effective noise power spectral density of a detector at different frequencies. The module contains both the LISA and TianQin sensitivity curves that can be tweaked by adjusting parameters such as the observation time, confusion noise and even the arm length. Additionally, this module allows the user to specify a custom detector sensitivity curve. We discuss the relevant equations in Sec.~\ref{sec:deriv_psd}.

@@

We developed \lw{} with an emphasis on increasing the efficiency of these computations in order to make large scale simulations more tractable. As a simple first step, we ensured that the entirety of \lw{} is vectorised and thus scales well with larger populations. In addition, we made a several specific optimisations to further increase the speed of calculations.

@@

Lastly, as a simple (yet thoroughly effective) optimisation, we interpolate (1) $g(n, e)$, which is a complicated combination of Bessel functions used in several calculations (see Eq.~\ref{eq:g(n,e)}) (2) the given sensitivity curve $S_{\rm n}(f)$. This significantly reduces the runtime of most strain and SNR calculations for large populations of sources.

\section{Derivations}\label{sec:derivations}
In this section, we present a derivation of the equations using in \lw{}. We emphasise that these are not new derivations, conversely, they are in fact given frequently in the literature. However, they are often incomplete, unclear or contain spurious constant factors that arise from invalid combinations of previous work. Here, we aim to present a clear, clean and concise explanation of the expressions we use in \lw{}.

A \docsIcon{} symbol in an equation will link to the relevant \lw{} documentation for the implementation of that equation, which additionally contains a link to exact code used to reproduce the equation. These derivations are also given in more detail in the \lw{} documentation \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{here}, where we show each of the intervening steps in more detail.

\subsection{Conversions and definitions {\normalfont (\texorpdfstring{\lwModLink{utils}}{utils})}}\label{sec:deriv_utils}
We start these derivations by defining some useful conversions and definitions. The chirp mass of a binary is the mass quantity measured by LISA and is given by
\begin{equation}{\lwFuncURL{utils}{chirp_mass}}
    \mathcal{M}_c \equiv \frac{(m_1 m_2)^{3/5}}{(m_1 + m_2)^{1/5}},
    \label{eq:chirpmass}
\end{equation}
where $m_1$ and $m_2$ are the primary and secondary masses of the double compact object.

It is often convenient to convert between orbital frequency, $f_{\rm orb}$, and the semi-major axis, $a$, of a binary and this can be accomplished with Kepler's third law.
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    a = \qty(\frac{G(m_1 + m_2)}{(2 \pi f_{\rm orb})^{2}})^{1/3},
    \label{eq:kepler3rd}
\end{equation}
or inversely
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    f_{\rm orb} = \frac{1}{2 \pi} \sqrt{\frac{G(m_1 + m_2)}{a^3}}.
    \label{eq:kepler3rd_reverse}
\end{equation}

For eccentric binaries, we need to consider all frequency harmonics of gravitational-wave emission. These are defined such that the $n^{\rm th}$ harmonic frequency is
\begin{equation}{}
    f_{n} \equiv n \cdot f_{\rm orb}.
    \label{eq:nth_freq}
\end{equation}
It will be important to know the relative gravitational-wave power radiated into the $n^{\rm th}$ harmonic for a binary with eccentricity $e$ for the following strain and SNR calculations. This is given by \citep[][Eq.\,20]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_g}}
    \begin{aligned}
        g(n, e) = \frac{n^{4}}{32} & \Big\{ \Big[ J_{n-2}(n e)-2 e J_{n-1}(n e)+\frac{2}{n} J_{n}(n e) \\
        &\left. +2 e J_{n+1}(n e)-J_{n+2}(n e)\Big]^{2}\right.\\
        &\left.\kern-\nulldelimiterspace\right.
            \!\!\begin{aligned}
            +\big(1-e^{2}\big)\Big[&J_{n-2}(n e)-2 J_{n}(n e) \\
            &+J_{n+2}(n e)\Big]^{2} \\
            \end{aligned}\\
        &+\frac{4}{3 n^{2}}\Big[J_{n}(n e)\Big]^{2}\Big\},
    \end{aligned}
    \label{eq:g(n,e)}
\end{equation}
where $J_{n}(v)$ is the Bessel function of the first kind. Thus, the \textit{sum} of $g(n, e)$ over all harmonics gives the factor by which the gravitational-wave emission is stronger for a binary of eccentricity $e$ over an equivalent circular binary. This enhancement factor is \citep[][Eq.\,17]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_f}}
    F(e) \equiv \sum_{n = 1}^{\infty} g(n, e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}}.
    \label{eq:eccentricity_enhancement_factor}
\end{equation}
A useful rule of thumb is that $F(0.5) \approx 5.0$, or in words, a binary with eccentricity $0.5$ loses energy to gravitational waves at 5 times the rate of an equivalent circular binary.

\subsection{Orbital evolution {\normalfont (\texorpdfstring{\lwModLink{evol}}{evol})}}\label{sec:deriv_evol}
\subsubsection{Circular binaries}
For a circular binary, the orbital evolution due to gravitational-wave emission can be calculated analytically, as the rate at which the separation of the binary shrinks is simply a function of its mass and the current separation \citep[][Eq.~5.6]{Peters+1964}
\begin{equation}{}
    \dv{a}{t}_{e = 0} = -\frac{\beta}{a^3},
\end{equation}
where the constant $\beta$ is defined as
\begin{equation}{\lwFuncURL{utils}{beta}}
    \beta(m_1, m_2) = \frac{64}{5} \frac{G^3}{c^5} m_1 m_2 (m_1 + m_2),
    \label{eq:beta_peters}
\end{equation}
where $m_1$ is the primary mass and $m_2$ is the secondary mass. This gives the semi-major axis of a circular binary as a function of time, $t$, as \citep[][Eq.~5.9]{Peters+1964}
\begin{equation}{\lwFuncURL{evol}{evol_circ}}
    a(t, m_1, m_2) = [a_0^4 - 4 t \beta(m_1, m_2)]^{1/4},
    \label{eq:a_over_time_circ}
\end{equation}
where $a_0$ is the initial semi-major axis. Moreover, we can solve for the merger time, the time until the binary will merge, by setting the final semi-major axis in Eq.~\ref{eq:a_over_time_circ}
\begin{equation}{\lwFuncURL{evol}{get_t_merge_circ}}
    t_{\rm merge, circ} = \frac{a_0^4}{4 \beta}.
    \label{eq:t_merge_circular}
\end{equation}

\subsubsection{Eccentric binaries}
The orbital evolution is more complex for eccentric binaries since the semi-major axis and eccentricity both evolve simultaneously and depend on one another. These equations cannot be solved analytically and require numerical integration. The semi major axis, $a$, and eccentricity, $e$, are related as \citep[][Eq.~5.11]{Peters+1964}
\begin{equation}{\lwFuncURL{utils}{get_a_from_ecc}}
    a(e) = c_0 \frac{e^{12/19}}{(1 - e^2)} \left(1 + \frac{121}{304} e^2\right)^{870/2299},
    \label{eq:a_from_e}
\end{equation}
where $c_0$ satisfies the initial conditions such that $a(e_0) = a_0$. The time derivative of the eccentricity, $e$, is given by
\begin{equation}{\lwFuncURL{evol}{de_dt}}
    \frac{\mathrm{d}e}{\mathrm{d}t} = -\frac{19}{12} \frac{\beta}{c_{0}^{4}} \frac{e^{-29 / 19}\left(1-e^{2}\right)^{3 / 2}}{\left[1+(121 / 304) e^{2}\right]^{\frac{1181}{2299}}},
    \label{eq:dedt}
\end{equation}
which we can integrate to find $e(t)$ and convert to $a(t)$ using Eq.~\ref{eq:a_from_e} (\docsLink{\lwFuncURL{evol}{evol_ecc}}).

Inverting this function and applying the fact that we know that $e \to 0$ when the binary merges gives the merger time \citep[][Eq.~5.14]{Peters+1964}
\begin{equation}{\lwFuncURL{evol}{get_t_merge_ecc}}
    t_{\rm merge} = \frac{12}{19} \frac{c_{0}^{4}}{\beta} \int_0^{e_0} \frac{\left[1+(121 / 304) e^{2}\right]^{\frac{1181}{2299}}}{e^{-29 / 19}\left(1-e^{2}\right)^{3 / 2}} \mathrm{d}e.
    \label{eq:t_merge_eccentric}
\end{equation}
For very small or very large eccentricities we approximate this integral using the following expressions (given in unlabelled equations after \citealp[][Eq.~5.14]{Peters+1964})
\begin{align}
    t_{\rm merge,\, e^2 \ll 1} &= \frac{c_0^4}{4 \beta} \cdot e_0^{48 / 19}, \\
    t_{\rm merge,\, (1 - e^2) \ll 1} &= \frac{768}{425} \frac{a_0^4}{4 \beta} (1 - e_0^2)^{7/2}.
\end{align}
\noindent The standard threshold employed by \lw{} for small eccentricities is $e=0.01$ and for large eccentricities is $e=0.99$, though we note that this can be customised by the user as well.

\subsection{Strains {\normalfont (\texorpdfstring{\lwModLink{strain}}{strain})}}\label{sec:deriv_strain}
The strength of a gravitational wave in a detector at any one moment is determined by the strain amplitude, $h_0$. However, for the LISA mission, the signal can be present in the detector for many years. This means that, the $n^{\rm th}$ harmonic of the binary will spend approximately $f_n / \dot{f}_n$ seconds (or $f_n^2 / \dot{f}_n$ cycles) in the vicinity of a frequency $f_n$ \citep{Finn+2000}. This leads to the signal `accumulating' at the frequency $f_n$.

Therefore, to account for the integration of the signal over the mission, we can introduce the `characteristic' strain amplitude of the $n^{\rm th}$ harmonic \citep[e.g][]{Finn+2000, Moore+2015}\footnote{Note that this is factor of 2 different from \citet{Finn+2000}. This is because the factor of 2 is already included in the \citet{Robson+2019} sensitivity curve and so is removed here.}
\begin{equation}{}
    h_{c, n}^2 = \qty(\frac{f_n^2}{\dot{f}_n}) h_{0, n}^2,
    \label{eq:strain-charstrain}
\end{equation}
where $h_{0, n}$ is the strain amplitude in the $n^{\rm th}$ harmonic.

@@

The characteristic strain from a stellar mass binary in the $n^{\rm th}$ harmonic is given by (e.g. \citealp[][Eq.\,56]{Barack+2004}; \citealp[][Eq.\,5.1]{ Flanagan+1998})
\begin{equation}{}
    h_{c,n}^2 = \frac{1}{(\pi D_L)^2} \qty( \frac{2 G}{c^3} \frac{\dot{E_n}}{\dot{f_n}} ),
    \label{eq:char_strain_dedf}
\end{equation}
where $D_L$ is the luminosity distance to the binary, $\dot{E}_n$ is the power radiated in the $n^{\rm th}$ harmonic and $\dot{f}_n$ is the rate of change of the $n^{\rm th}$ harmonic frequency.

>>>>

\title{LEGWORK: A python package for computing the evolution and detectability \\ of stellar-origin gravitational-wave sources with space-based detectors}

@@

\begin{abstract}
% Original
%    We present \lw{}, a Python package for making predictions about stellar-origin gravitational wave sources and their detectability in LISA and other space-based gravitational wave detectors. \lw{} can be used to evolve sources over time, calculate gravitational wave strains, compute signal-to-noise ratios and visualise the results. In this paper we give an overview of the structure and functionality of the package, as well as highlight several optimisations that make it particularly efficient for use in population studies which can contain tens-of-millions of sources. We demonstrate the various capabilities of \lw{} through a series of example use cases. These begin with a basic example of calculating the signal-to-noise ratio of a source but also include using \lw{} to calculate horizon distances and make comparisons between gravitational-wave detectors. We additionally explain in detail the derivation of the equations behind \lw{} in order to clarify some conflicts currently present in the literature.

% -------------

% Selma version
% We present \lw{} (LISA Evolution and Gravitational Wave Orbit Kit) an open-source Python package developed to aid studies of stellar-origin gravitational wave sources detectable by space-based gravitational wave detectors such as LISA. 
% %
% This package can be used to follow the gravitational wave driven inspiral of a compact binary consisting of, for example, white dwarfs, neutron stars and/or stellar-mass black holes. We calculate the gravitational wave strains using the post-Newtonian approximations and estimate the signal-to-noise ratio for a given detector sensitivity. We optimised \lw{} to make it feasible to make predictions for populations of tens-of-millions of sources. In principle, \lw{} can be used for any systems with a user-specified orbital evolution, such as those affected by a third object or gas drag. 
% %
% This paper describes the package and presents several use cases. We provide the derivations behind the expressions and we identify and clarify some discrepancies in the expressions presented in earlier literature. We hope that \lw{} will enable and accelerate future studies triggered by the rapidly growing interest in gravitational wave sources.

We present \lw{} (LISA Evolution and Gravitational Wave Orbit Kit), an open-source Python package for making predictions about stellar-origin gravitational wave sources and their detectability in LISA or other space-based gravitational wave detectors. \lw{} can be used to evolve the orbits of sources due to gravitational wave emission, calculate gravitational wave strains (using post-Newtonian approximations), compute signal-to-noise ratios and visualise the results. It can be applied to a variety of potential sources, including binaries consisting of white dwarfs, neutron stars and black holes. Although we focus on double compact objects, in principle \lw{} can be used for any system with a user-specified orbital evolution, such as those affected by a third object or gas drag. We optimised the package to make it efficient for use in population studies which can contain tens-of-millions of sources. This paper describes the package and presents several potential use cases. We explain in detail the derivations of the expressions behind the package as well as identify and clarify some discrepancies currently present in the literature. We hope that \lw{} will enable and accelerate future studies triggered by the rapidly growing interest in gravitational wave sources.
\end{abstract}

\keywords{open source software, gravitational waves, gravitational wave detectors, compact objects, orbital evolution, white dwarfs, neutron stars, stellar mass black holes }

@@

The planned space-based gravitational wave detector LISA (Laser Interferometer Space Antenna) will present an entirely new view of gravitational waves by focusing on lower frequencies ($10^{-5} < f / \unit{Hz} < 10^{-1}$) than ground-based detectors. This will enable the study of many new source classes including mergers of supermassive black holes \citep[e.g.][]{Begelman+1980, Klein+2016, Bellovary2019}, extreme mass ratio inspirals \citep[e.g.,][]{Berti2006, Barack2007, Babak2017, Moore2017}, and cosmological GW backgrounds \citep[e.g.,][]{Bartolo2016, Caprini2016, Caldwell2019}. However, this frequency regime is also of interest for the detection of local stellar-mass binaries during their inspiral phase. LISA is expected to detect Galactic stellar-origin binaries containing combinations of white dwarfs, neutron stars, and black holes, ranging from the numerous double white dwarf population, to the rare but loud double black hole population.

The potential to detect stellar-origin sources with LISA has been studied in the past \citep[e.g.][]{Nelemans+2001, Liu+2009, Liu+2014, Ruiter+2010, Belczynski+2010, Yu+2010, Nissanke+2012}. More recently, the direct detection of gravitational waves with ground-based detectors has led to renewed interest in this topic with many recent papers addressing the issue \citep[e.g.][]{Korol+2017, Korol+2018, Korol+2019, Korol+2020, Christian+2017, Kremer+2017, Kremer+2018, Lamberts+2018, Lamberts+2019, Fang+2019, Andrews+2020, Lau+2020, Breivik+2020, Breivik+2020a, Roebber+2020, Chen+2020, Sesana+2020, Shao+2021}.

@@

\lw{} is an open-source Python package designed to streamline the process of making predictions of LISA detection rates for stellar-origin binaries such that it is as fast, reliable and simple as possible. This goal makes \lw{} unique among other implementations of gravitational-wave tools in the literature, which focus on a more broad coverage of the gravitational-wave spectrum and source classes, rather than an optimised approach for certain sources. \citep[e.g.][]{Moore+2015,Yi2021}. With \lw{} one can evolve the orbits of a binary or a collection of binaries and calculate their strain amplitudes for any range of frequency harmonics. One can compute the sensitivity curve for LISA or other future gravitational wave detectors (e.g. TianQin's curve, or that of a custom instrument) and use it to compute the signal-to-noise ratio of a collection of sources. Furthermore, \lw{} provides tools to visualise all of the results with easy-to-use plotting functions. Finally, \lw{} is fully tested to check for consistency in the derivations described below.

Specifically, we implement the post-Newtonian expressions by \citet{Peters+1963} and \citet{Peters+1964} for the evolution of binary orbits due to the emission of gravitational waves, equations for the strain amplitudes and signal-to-noise ratios of binaries from various papers \citep[e.g.][]{Flanagan+1998, Finn+2000, Cornish2003, Barack+2004, Moore+2015} and approximations for the LISA and TianQin sensitivity curves given in \citet{Robson+2019} and \citet{Huang+2020} respectively.

@@

\lw{} has been adopted in multiple studies. Wagg et al.\ (in prep) investigates several populations of potential LISA sources (double black holes, black hole neutron stars and double neutron stars). They use \lw{} to evolve each synthesised source over the age of the Milky Way, make predictions about the LISA detectable population and explore how it varies with different binary physics assumptions. Additionally, Thiele et al.\ (in prep) examines the implications of assuming a metallicity-dependent binary fraction for the formation of close double white dwarfs (WDWDs) in the LISA frequency band. They use \lw{} to calculate the SNR of WDWDs and, in particular, apply a custom noise model that combines the LISA sensitivity curve from \citet{Robson+2019} with a new fit for the Galactic confusion noise based on their WDWD population.

@@

The \lw{} package is composed of seven modules that each focus on a particular aspect of calculations useful for gravitational wave sources that are detectable by space-based detectors. In Fig.~\ref{fig:package_overview}, we illustrate the general structure of the package with each of its modules. The \texttt{source} module is the central module of the package and provides an simple interface to the functions in the rest of the modules. For more complex analyses, users may want to interact directly with individual modules, particularly those in the top row of Fig.~\ref{fig:package_overview} as they comprise the core functionality of \lw{}. Below we explain the capabilities of each module in detail.

@@

@@

\lwModLink{strain} contains two functions that compute a binary's gravitational wave strain and characteristic strain amplitude respectively. Each of these functions is capable of computing the strain for an array of binaries at any number of timesteps and evaluated at any number of frequency harmonics. We discuss the relevant equations in Sec.~\ref{sec:deriv_strain}.

\lwModLink{psd} is used for evaluating the effective noise power spectral density of a detector at different frequencies. The module currently contains the LISA and TianQin sensitivity curves that can be tweaked by adjusting parameters such as the observation time, confusion noise and even the arm length. Additionally, this module allows the user to specify a custom detector sensitivity curve. We discuss the relevant equations in Sec.~\ref{sec:deriv_psd}.

@@

We developed \lw{} with an emphasis on increasing the efficiency of these computations in order to make simulations of large populations of systems tractable. We ensured that the entirety of \lw{} is vectorised and thus scales well with larger populations. In addition, we made a several specific optimisations to further increase the speed of calculations.

Firstly, we found that the runtime of calculating strains and SNRs for large populations of sources was mainly limited by the computation of (1) the relative gravitational-wave power in each harmonic for eccentric systems (see Eq.~\ref{eq:g(n,e)}) (2) the sensitivity curve of the given detector. Therefore, in order to significantly reduce the runtime of strain and SNR computations, by default \lw{} automatically interpolates these functions upon instantiation of any \texttt{Source} class with a large number of sources. Thereafter all functions use the tabulated values instead of calculating them exactly.

@@

\section{Derivations}\label{sec:derivations}
In this section, we present a derivation of the equations used in \lw{}. We emphasise that these are not new derivations, conversely, they are in fact given frequently in the literature. However, they are often incomplete, unclear and, in some cases, contain spurious constant factors that arise from invalid combinations of previous work. Here, we aim to present a clear, clean and concise explanation of the expressions we use in \lw{}.

A \docsIcon{} symbol in an equation directly links to the relevant online \lw{} documentation for the implementation of that equation, which additionally contains a link to exact code used to reproduce the equation. These derivations are also given in more detail in the \lw{} documentation \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{here}, where we show each of the intervening steps in more detail.

\subsection{Conversions and definitions {\normalfont (\texorpdfstring{\lwModLink{utils}}{utils})}}\label{sec:deriv_utils}
We start these derivations by defining some useful conversions and definitions. The chirp mass of a binary is the mass quantity measured by LISA and is given by
\begin{equation}{\lwFuncURL{utils}{chirp_mass}}
    \mathcal{M}_c \equiv \frac{(m_1 m_2)^{3/5}}{(m_1 + m_2)^{1/5}},
    \label{eq:chirpmass}
\end{equation}
where $m_1$ and $m_2$ are the primary and secondary masses of the binary.

It is often convenient to convert between orbital frequency, $f_{\rm orb}$, and the semi-major axis, $a$, of a binary and this can be accomplished with Kepler's third law.
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    a = \qty(\frac{G(m_1 + m_2)}{(2 \pi f_{\rm orb})^{2}})^{1/3},
    \label{eq:kepler3rd}
\end{equation}
where $G$ is the gravitational constant. Inversely we could write
\begin{equation}{\lwFuncURL{utils}{get_a_from_f_orb}}
    f_{\rm orb} = \frac{1}{2 \pi} \sqrt{\frac{G(m_1 + m_2)}{a^3}}.
    \label{eq:kepler3rd_reverse}
\end{equation}

For circular binaries, gravitational-wave emission occurs at twice the orbital frequency ($f_{\rm GW} = 2 f_{\rm orb}$). However, for eccentric binaries, we need to consider all frequency harmonics of gravitational-wave emission. These are defined such that the $n^{\rm th}$ harmonic frequency is
\begin{equation}{}
    f_{n} \equiv n \cdot f_{\rm orb}.
    \label{eq:nth_freq}
\end{equation}
It will be important to know the relative gravitational-wave power radiated into the $n^{\rm th}$ harmonic for a binary with eccentricity $e$ for the following strain and SNR calculations. This is given by \citet[][Eq.\,20]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_g}}
    \begin{aligned}
        g(n, e) = \frac{n^{4}}{32} & \Big\{ \Big[ J_{n-2}(n e)-2 e J_{n-1}(n e)+\frac{2}{n} J_{n}(n e) \\
        &\left. +2 e J_{n+1}(n e)-J_{n+2}(n e)\Big]^{2}\right.\\
        &\left.\kern-\nulldelimiterspace\right.
            \!\!\begin{aligned}
            +\big(1-e^{2}\big)\Big[&J_{n-2}(n e)-2 J_{n}(n e) \\
            &+J_{n+2}(n e)\Big]^{2}
            \end{aligned}\\
        &+\frac{4}{3 n^{2}}\Big[J_{n}(n e)\Big]^{2}\Big\},
    \end{aligned}
    \label{eq:g(n,e)}
\end{equation}
where $J_{n}(v)$ is the Bessel function of the first kind. Thus, the \textit{sum} of $g(n, e)$ over all harmonics gives the factor by which the gravitational-wave emission is stronger for a binary of eccentricity $e$ over an equivalent circular binary. This enhancement factor is \citep[][Eq.\,17]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_f}}
    F(e) \equiv \sum_{n = 1}^{\infty} g(n, e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}}.
    \label{eq:eccentricity_enhancement_factor}
\end{equation}
A useful rule of thumb is that $F(0.5) \approx 5.0$, or in words, a binary with eccentricity $0.5$ loses energy to gravitational waves at 5 times the rate of an equivalent circular binary.

\subsection{Orbital evolution {\normalfont (\texorpdfstring{\lwModLink{evol}}{evol})}}\label{sec:deriv_evol}
\subsubsection{Circular binaries}
For a circular binary, the orbital evolution due to gravitational-wave emission can be calculated analytically, as the rate at which the separation of the binary shrinks is simply a function of its mass and the current separation \citep[][Eq.~5.6]{Peters+1964}
\begin{equation}{}
    \dv{a}{t}_{e = 0} = -\frac{\beta}{a^3},
\end{equation}
where the constant $\beta$ is defined as
\begin{equation}{\lwFuncURL{utils}{beta}}
    \beta(m_1, m_2) \equiv \frac{64}{5} \frac{G^3}{c^5} m_1 m_2 (m_1 + m_2),
    \label{eq:beta_peters}
\end{equation}
where $c$ is the speed of light, $m_1$ is the primary mass and $m_2$ is the secondary mass. This gives the semi-major axis of a circular binary as a function of time, $t$, as \citep[][Eq.~5.9]{Peters+1964}
\begin{equation}{\lwFuncURL{evol}{evol_circ}}
    a(t, m_1, m_2) = [a_0^4 - 4 t \beta(m_1, m_2)]^{1/4},
    \label{eq:a_over_time_circ}
\end{equation}
where $a_0$ is the initial semi-major axis. Moreover, we can solve for the merger time, the time until the binary will merge, by setting the final semi-major axis in Eq.~\ref{eq:a_over_time_circ}
\begin{equation}{\lwFuncURL{evol}{get_t_merge_circ}}
    t_{\rm merge, circ} = \frac{a_0^4}{4 \beta}.
    \label{eq:t_merge_circular}
\end{equation}

\subsubsection{Eccentric binaries}
The orbital evolution is more complex for eccentric binaries since the semi-major axis and eccentricity both evolve simultaneously and depend on one another. The final expression cannot be solved analytically and require numerical integration. The semi major axis, $a$, and eccentricity, $e$, are related as \citep[][Eq.~5.11]{Peters+1964}
\begin{equation}{\lwFuncURL{utils}{get_a_from_ecc}}
    a(e) = c_0 \frac{e^{12/19}}{(1 - e^2)} \left(1 + \frac{121}{304} e^2\right)^{870/2299},
    \label{eq:a_from_e}
\end{equation}
where $c_0$ satisfies the initial conditions such that $a(e_0) = a_0$. The time derivative of the eccentricity, $e$, is given by
\begin{equation}{\lwFuncURL{evol}{de_dt}}
    \frac{\mathrm{d}e}{\mathrm{d}t} = -\frac{19}{12} \frac{\beta}{c_{0}^{4}} \frac{e^{-29 / 19}\left(1-e^{2}\right)^{3 / 2}}{\left[1+(121 / 304) e^{2}\right]^{\frac{1181}{2299}}},
    \label{eq:dedt}
\end{equation}
which we can integrate to find $e(t)$ and convert to $a(t)$ using Eq.~\ref{eq:a_from_e} (\docsLink{\lwFuncURL{evol}{evol_ecc}}).

Inverting this function and applying the fact that we know that $e \to 0$ when the binary merges gives the merger time \citep[][Eq.~5.14]{Peters+1964}
\begin{equation}{\lwFuncURL{evol}{get_t_merge_ecc}}
    t_{\rm merge} = \frac{12}{19} \frac{c_{0}^{4}}{\beta} \int_0^{e_0} \frac{\left[1+(121 / 304) e^{2}\right]^{\frac{1181}{2299}}}{e^{-29 / 19}\left(1-e^{2}\right)^{3 / 2}} \mathrm{d}e.
    \label{eq:t_merge_eccentric}
\end{equation}
For very small or very large eccentricities we approximate this integral using the following expressions (given in unlabelled equations after \citealp[][Eq.~5.14]{Peters+1964})
\begin{align}
    t_{\rm merge,\, e^2 \ll 1} &= \frac{c_0^4}{4 \beta} \cdot e_0^{48 / 19}, \\
    t_{\rm merge,\, (1 - e^2) \ll 1} &= \frac{768}{425} \frac{a_0^4}{4 \beta} (1 - e_0^2)^{7/2}.
\end{align}
\noindent The standard threshold employed by \lw{} for small eccentricities is $e = 0.15$ and for large eccentricities is $e=0.9999$ (as this approximates $t_{\rm merge}$ with an error below roughly 2\%), though we note that this can be customised by the user if desired.

In addition, we implement the fit to Eq.~\ref{eq:t_merge_eccentric} from \citet{Mandel+2021} that approximates the merger time as
\begin{equation}{\lwFuncURL{evol}{t_merge_mandel_fit}}
    \begin{split}
        t_{\rm merge} \approx& t_{\rm merge, circ} (1 - e^2_0)^{7/2} \\
    &\times(1 + 0.27 e_0^{10} + 0.33 e^{20}_0 + 0.2 e_0^{1000} ),
    \end{split}
\end{equation}
which gives $t_{\rm merge}$ with an error below 3\% for eccentricities below $0.9999$. We additionally add a rudimentary polynomial fit to further reduce this error to below $0.5\%$. The user may specify whether to use this fit or perform the full integral when calculating merger times in \lw{}.

\subsection{Strains {\normalfont (\texorpdfstring{\lwModLink{strain}}{strain})}}\label{sec:deriv_strain}
The strength of a gravitational wave in a detector at any one moment is determined by the strain amplitude, $h_0$. However, for stellar-origin sources at mHz frequencies, the signal can be present in the detector for many years. This means that, the $n^{\rm th}$ harmonic of the binary will spend approximately $f_n / \dot{f}_n$ seconds (or $f_n^2 / \dot{f}_n$ cycles) in the vicinity of a frequency $f_n$ \citep{Finn+2000}. This leads to the signal `accumulating' at the frequency $f_n$.

Therefore, to account for the integration of the signal over the mission, we instead use the `characteristic' strain amplitude of the $n^{\rm th}$ harmonic \citep[e.g][]{Finn+2000, Moore+2015}\footnote{Note that this is factor of 2 different from \citet{Finn+2000}. This is because the factor of 2 is already included in the \citet{Robson+2019} sensitivity curve and so is removed here.}
\begin{equation}{}
    h_{c, n}^2 = \qty(\frac{f_n^2}{\dot{f}_n}) h_{0, n}^2,
    \label{eq:strain-charstrain}
\end{equation}
where $h_{0, n}$ is the strain amplitude in the $n^{\rm th}$ harmonic.

@@

The characteristic strain from a stellar mass binary in the $n^{\rm th}$ harmonic is given by (e.g. \citealp[][Eq.\,56]{Barack+2004}; \citealp[][Eq.\,5.1]{ Flanagan+1998})
\begin{equation}{}
    h_{c,n}^2 = \frac{1}{(\pi D_L)^2} \qty( \frac{2 G}{c^3} \frac{\dot{E_n}}{\dot{f_n}} ),
    \label{eq:char_strain_dedf}
\end{equation}
where $D_L$ is the luminosity distance to the source (note that for Milky Way sources this is simply the distance to the source), $\dot{E}_n$ is the power radiated in the $n^{\rm th}$ harmonic and $\dot{f}_n$ is the rate of change of the $n^{\rm th}$ harmonic frequency.
